Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-03-11 bbbq_article.tex
Plot number of SNPs per TMP, fix #143
Processed feedback Frans, thanks @fransbianchi. Fix #188

Cytolytic T cell responses are predicted to be biased towards membrane proteins. 
Because the peptide-binding grooves of most haplotypes 
of histocompatibility complex class I (MHC-I) are relatively hydrophobic, 
peptide fragments derived from transmembrane helices (TMHs) 
are predicted to be presented more often than expected based on their abundance.
However, the physiological reason of why membrane proteins might be 
over-presented is unclear
In this study, we show the TMHs are evolutionarily more conserved, 
as relatively less single nucleotide polymorphisms (SNPs) 
are present in TMH-coding chromosomal regions 
compared to regions coding for extracellular and cytosolic protein regions. 
Moreover, we show that the over-presentation of TMH-derived peptides is 
likely general, as it is predicted for diverse microbial pathogens 
and for both MHC-I and MHC-II.
Thus, our findings suggest that T cells might respond 
more to membrane proteins, because these are evolutionary more conserved.
We speculate that TMHs might be less prone of the escape mutations 
that enable pathogens to evade T cell responses.

@@

Our immune system fights microbial pathogens, 
such as fungi, bacteria or viruses. 
An important part of the acquired immune response, 
that develops specialized and more effective recognition of pathogens, 
are T cells which recognize peptides derived from the pathogens 
by use of Major Histocompatibility Complexes (MHC).

@@

Any individual's immune system detects only a fraction of all possible
peptide fragments.
For humans, the MHC proteins are encoded by the
HLA (Human Leukocyte Antigens) genes.
There are three genes encoding for MHC-I, which are HLA-A, HLA-B and HLA-C,
for MHC-II there are three major genes, which are HLA-DR, HLA-DQ, HLA-DP.
Each MHC complex can only bind a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition}).
The HLA region of humans is highly polymorphic, with hundreds 
to thousands of different alleles, 
and each different HLA gene is called 
an MHC haplotype (\cite{marsh2010nomenclature}).

@@

Because of these multiple and highly polymorphic MHC genes,
the number of pathogenic peptides that can be detected is increased,
as a wide variety of MHCs will be expressed,
each presenting their own subset of epitopes.
It is believed that this improves immunity at the population level, 
as mutations in a protein that disrupt a particular MHC presentation, 
so-called escape mutations, 
will not affect MHC presentation for all haplotypes (\cite{sommer2005importance}).

@@

Much studies are aimed at determining which peptides are presented in MHC 
and will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development 
of very reliable prediction algorithms 
that allows for in silico predictions 
of the binding affinities of peptides (\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}).

@@

This over-presentation is entirely explained by the fact 
that the peptide-binding groove of most MHC-I haplotypes 
is relatively hydrophobic, 
and therefore hydrophobic peptides have a higher affinity 
than hydrophilic peptides.

Transmembrane helices (TMHs) are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes, 
and they consist of an alpha helix of usually 23 amino acids. 
TMHs are common structures in the proteins of humans and pathogens. 
Different TMH prediction tools estimate
that 15-39% of the proteins in humans 
contain at least one TMH (\cite{ahram2006estimation}). 
TMHs can also be predicted from a protein sequence 
with high accuracy by bioinformatics approaches (\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}).

However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble protein regions is unknown. 
One possibility is that the presentation of 
hydrophobic/TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutation. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to this, there are constraints in the types of amino acids 
that can be present in the TMH (\cite{jones1994model}). 
Therefore, the TMHs of pathogens 
might have a lower chance to develop an escape mutations, 
as many mutations will result in a dysfunctional TMH 
and render the protein inactive.

@@

>>>>

Cytolytic T cell responses are predicted to be biased towards membrane proteins. 
Because the peptide-binding grooves of most haplotypes 
of histocompatibility complex class I (MHC-I) are relatively hydrophobic, 
peptide fragments derived from transmembrane helices (TMHs) 
are predicted to be presented more often than expected 
based on their abundance in the proteome.
We show that the over-presentation of TMH-derived peptides is likely general, 
as it is predicted for diverse microbiota and viruses 
and for both MHC-I and MHC-II
However, the physiological reason of why membrane proteins might be 
over-presented is unclear
In this study, we show the TMHs are evolutionarily more conserved, 
as relatively less single nucleotide polymorphisms (SNPs) 
are present in TMH-coding chromosomal regions 
compared to regions coding for extracellular and cytosolic protein regions. 
Thus, our findings suggest that T cells might respond 
more to membrane proteins, because these are evolutionary more conserved.
We speculate that TMHs therefor might be less prone to escape mutations 
that enable pathogens to evade T cell responses.

@@

Our immune system fights diseases and infections from pathogens, 
such as fungi, bacteria or viruses. 
An important part of the acquired immune response, 
that develops specialized and more effective recognition of pathogens, 
are T cells which recognize peptides derived from 
exogenous antigens presented on Major Histocompatibility Complexes (MHC).

@@

Any individual's immune system detects only a fraction of all possible
peptide fragments.
For humans, the MHC proteins are encoded by the
HLA (Human Leukocyte Antigens) genes.
There are three genes encoding for MHC-I, which are HLA-A, HLA-B and HLA-C,
for MHC-II there are three major genes, which are HLA-DR, HLA-DQ, HLA-DP.
Each MHC complex can only bind a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition})
\frans{Is dit zo?} 
\richel{Volgens de (weggegevallen) referentie naar Lund et al., 2004, 'Definition of supertypes for {HLA} molecules using clustering of specificity matrices' wel}.
The HLA region of humans is highly polymorphic, with hundreds 
to thousands of different alleles, 
and each different HLA gene is called 
an MHC haplotype (\cite{marsh2010nomenclature}).

@@

Because of these multiple and highly polymorphic MHC genes,
the number of pathogenic peptides that can be detected is increased,
as a wide variety of MHCs: 
6 MHC-I and 6-8 MHC-II molecules are expressed per individual
each presenting their own subset of epitopes.
It is believed that this improves immunity at the population level, 
as mutations in a protein that disrupt a particular MHC presentation, 
so-called escape mutations, 
will not affect MHC presentation for all haplotypes (\cite{sommer2005importance}).

@@

Many studies are aimed at determining which peptides are presented in MHC 
and will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development 
of very reliable prediction algorithms 
that allows for in silico predictions 
of the binding affinities of peptides (\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}).

@@

This over-presentation is attributed to the fact 
that the peptide-binding groove of most MHC-I haplotypes 
is relatively hydrophobic, 
and therefore hydrophobic peptides have a higher affinity 
than hydrophilic peptides.

Transmembrane helices (TMHs) are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix on average 23 amino acids in length. 
TMHs are common structures in the proteins of humans and pathogens. 
Different TMH prediction tools estimate
that 15-39\% of the proteins in humans 
contain at least one TMH (\cite{ahram2006estimation}).
TMHs can also be predicted from a protein sequence 
with high accuracy by bioinformatics approaches (\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}).

However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble protein regions is unknown. 
We hypothesized is that the presentation of 
hydrophobic/TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutation. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to these requirements, 
the amino acids genuinely present in TMH 
are limited to the ones with hydrophobic side chains (\cite{jones1994model}).
\frans{Analyzing the occurrence of single nucleotide polymorphisms that were predicted to be inside ...}

Therefore, the TMHs of pathogens 
might have a lower chance to develop an escape mutations, 
as many mutations will result in a dysfunctional TMH 
and render the protein inactive.

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_n_snps_per_tmp.png}
  \caption{
    Histogram of the number of SNPs per trans-membrane protein.
  }
  \label{fig:n_tmhs_per_protein}
\end{figure}
